Technology for application family creation based on domain analysis

### 2016-04-03 2016-domainAnalysis/domainAnalysis.tex
Some more references

The next thing we need is to generate actual code on target textual language that will call feature library and glue features together. For this we shall return to metamodel level and define generation rules for metamodel. This step is performed only once for a given domain after the feature library and metamodel are finished, and then the same generator is used for each application created by using of the technology. Recommendations for development of domain-specific generator are well-known in DSM literature (for example, []): it is the best to write first application by hand, then draw a model that is supposed to be generated into this application, then find the places in handwritten application that shall be parameterised by information from model and let the generator replace such handwritten parts with data from model. This process is continued until handwritten application becomes a template that is filled by generator with information taken from model. Handwritten application and, consequently, a generator shall extensively use feature library to minimize the amount of code that is generated directly, in ideal case generator shall produce merely a glue code that binds components from feature library together.

After all steps above are finished we have feature library, visual editor for simple domain-specific language that allows to describe how features are combined and configured in a concrete application, and a generator that automatically produces complete application by a model in domain-specific language using feature library as domain-specific runtime[]. Now we may create as many applications as we wish by just drawing models and automatically generate complete executable code. Theoretically. Of course, in practise there will always be a need to modify feature diagram, to extend feature library and, consequently, domain-specific language metamodel, modify generator and even to make some changes in generated code, there is no silver bullet. But we believe that our approach can provide better separation of concerns, provides better utilization of domain experts knowledge and expertise among a team. Summary of a process described above and relation between various tools and roles of developers is provided on Fig.~\ref{image:process}.

This approach was implemented in a technology based on QReal DSM platform. QReal became an enabler technology because it provides easy and effective way to create visual editor for domain-specific languages that allows to create fully functional editor in less than an hour. It has visual metaeditor, visual constraints definition tool, visual shape editor and a C++ library that allows to quickly specify generation rules. Feature diagram editor and generator that creates metamodel by feature diagrams were both implemented as plugins to QReal core. Note that feature diagram language is itself domain-specific language for the domain of domain analysis, so it was implemented using QReal metaeditor. The same metaeditor (including shape editor and constraints editor) is then used to tailor the generated metamodel of domain-specific language. Then metaeditor generator is used to generate yet another plugin to QReal that provides visual editor for created language. Then the generator is implemented by hand on C++ with Qt library[] using generator creation library included in QReal. Then it is possible to create special distribution of QReal (using Qt Installer framework) that includes only QReal core, editors for feature diagrams (at this point they are needed only as reference) and domain-specific language, generator and feature library, thus forming a complete technology that can be used to generate target applications.

>>>>

The next thing we need is to generate actual code on target textual language that will call feature library and glue features together. For this we shall return to metamodel level and define generation rules for metamodel. This step is performed only once for a given domain after the feature library and metamodel are finished, and then the same generator is used for each application created by using of the technology. Recommendations for development of domain-specific generator are well-known in DSM literature (for example,~\cite{kelly2008domain}): it is the best to write first application by hand, then draw a model that is supposed to be generated into this application, then find the places in handwritten application that shall be parameterised by information from model and let the generator replace such handwritten parts with data from model. This process is continued until handwritten application becomes a template that is filled by generator with information taken from model. Handwritten application and, consequently, a generator shall extensively use feature library to minimize the amount of code that is generated directly, in ideal case generator shall produce merely a glue code that binds components from feature library together.

After all steps above are finished we have feature library, visual editor for simple domain-specific language that allows to describe how features are combined and configured in a concrete application, and a generator that automatically produces complete application by a model in domain-specific language using feature library as domain-specific runtime~\cite{kelly2008domain}. Now we may create as many applications as we wish by just drawing models and automatically generate complete executable code. Theoretically. Of course, in practise there will always be a need to modify feature diagram, to extend feature library and, consequently, domain-specific language metamodel, modify generator and even to make some changes in generated code, there is no silver bullet. But we believe that our approach can provide better separation of concerns, provides better utilization of domain experts knowledge and expertise among a team. Summary of a process described above and relation between various tools and roles of developers is provided on Fig.~\ref{image:process}.

This approach was implemented in a technology based on QReal DSM platform. QReal became an enabler technology because it provides easy and effective way to create visual editor for domain-specific languages that allows to create fully functional editor in less than an hour. It has visual metaeditor, visual constraints definition tool, visual shape editor and a C++ library that allows to quickly specify generation rules. Feature diagram editor and generator that creates metamodel by feature diagrams were both implemented as plugins to QReal core. Note that feature diagram language is itself domain-specific language for the domain of domain analysis, so it was implemented using QReal metaeditor. The same metaeditor (including shape editor and constraints editor) is then used to tailor the generated metamodel of domain-specific language. Then metaeditor generator is used to generate yet another plugin to QReal that provides visual editor for created language. Then the generator is implemented by hand on C++ with Qt library\footnote{Qt library home page, URL: http://www.qt.io/ (03.04.2016)} using generator creation library included in QReal. Then it is possible to create special distribution of QReal (using Qt Installer framework\footnote{Qt Installer Framework home page, URL: https://wiki.qt.io/Qt-Installer-Framework (03.04.2016)}) that includes only QReal core, editors for feature diagrams (at this point they are needed only as reference) and domain-specific language, generator and feature library, thus forming a complete technology that can be used to generate target applications.
